---
output:
  word_document: default
  html_document: default
---
Problem 1: A university has the following academic ranks for tenured/tenure-track faculty: assistant professor, associate
professor, and professor. The first two ranks are tenure-track, so only professors have tenure. A tenure-track faculty member
may be discharged, remain at his or her present rank, or be promoted to a higher rank. Discharged faculty are never rehired.
Only tenure-track faculty members can be promoted or discharged. A professor may remain as his or her rank or retire. All
changes happen at the end of an academic year.


a. The career path of a faculty member can be modeled using a Markov chain with 5 states. Define the state variable, the
state space, and give the index set for the Markov chain.
1.Discharged 
2.Assistant Professor
3.Associate Professor
4.Professor 
5.Retire

State Space {1,2,3,4,5}

b.The promotion and discharge probabilities for assistant professors are 0.3 and 0.2, respectively. The promotion and
discharge probabilities for associate professors are 0.2 and 0.15, respectively. On average, 10% of professors retire
each year. Construct a transition matrix and transition diagram for the Markov chain.

```{r}
library(diagram)
library(markovchain)

Discharged <- c(1,0,0,0,0)
A_Prof<-c(0.2,0.5,0.3,0,0)
As_Prof<- c(0.15,0,0.65,0.2,0)
Prof<-c(0,0,0,0.9,0.1)
Retired <- c(0,0,0,0,1)

x <- cbind(Discharged,A_Prof,As_Prof,Prof,Retired)
colnames(x)<- c("Discharged","A_Prof","As_Prof","Prof","Retired")
transition_matrix<- matrix(x, nrow =5 ,byrow=TRUE)


dtmcA <- new("markovchain",transitionMatrix=transition_matrix,states=c("Discharged","A_Prof","As_Prof","Prof","Retired"), name="MarkovChain A") 
dtmcA

plot(dtmcA)


```
c.If the university’s faculty consists of 45% assistant professors, 30% associate professors, and 25% professors at the
end of this academic year, what should the distribution of faculty be at the end of the next academic year? Express
answers to the nearest 0.001.

```{r}
#current distribution
c <- c(0,0.45,0.3,0.25,0)
#distribution by year end
y<- c%*%transition_matrix
colnames(y) <- c("Discharged","Assistant_Professor","Associate_Professor","Professor","Retired")
y
```

Problem 2: Consider a timber growth model used in forest management. The states are size categories, measured by tree
diameter. Every three years, all the trees in a stand were measured, and from that data the following transition matrix was
determined. Units are in inches.

a. Define the state space for the Markov chain and draw the transition diagram.
State Space 
1. 0-1
2. 1-3
3. 3-7
4 8

State Space {1,2,3,4}
```{r}

library(diagram)
library(markovchain)

size_1 <- c(0.92,0.08,0,0)
size_2 <- c(0,0.97,0.03,0)
size_3 <- c(0,0,0.98,0.02)
size_4 <- c(0,0,0,1)

x <- cbind(size_1,size_2,size_3,size_4)
colnames(x)<- c("0-1","1-3","3-7","8")
rownames(x)<- c("0-1","1-3","3-7","8")
transition_matrix_2<- matrix(x, nrow =4 ,byrow=TRUE)
dtmcA <- new("markovchain",transitionMatrix=transition_matrix_2,states=c("0-1","1-3","3-7","8"), name="MarkovChain A") 
dtmcA
plot(dtmcA)

```
b. Note that the time for each transition is 3 years. This means that the index set is T = {0;1;2;3;4; : }and each number
represents a 3-year period. So X1 is the state variable representing the size of a tree after 1 transition which is 3
years. Similarly, X2 takes on the number of the state representing tree size after 6 years, or two transitions. Find the
percentage of the current trees in the stand that are currently 0-1 inches in diameters which, after 15 years will be:
i. 1-3 inches in diameter  - 31.9%
ii. 3-8 inches in diameter  -2.1%
iii. more than 8 inches in diameter -0.0449%
```{r}
library(matrixcalc)
#Each tranistion is tree years, so to get to 15,is 5 steps 
P_5 <- matrix.power(transition_matrix_2,5)
P_5
inch_1_3 <-P_5[1,2]
inch_1_3 
inch_3_8 <-P_5[1,3]
inch_3_8
inch_8  <- P_5[1,4]
inch_8 
```
c.Consider the current population of 0-1 inch diameter trees. Using the Markov model, find the percentage that will be
less than 1 inch in diameter after 30 years.- 43.4%
```{r}
#Each tranistion is tree years, so to get to 30,is 10 steps 
P_10 <- matrix.power(transition_matrix_2,10)
P_10
inch_0_1 <-P_10[1,1]
inch_0_1
```
d.Suppose that of the trees currently in the stand, 50% have less than 1 inch diameter, 28% have 1-3 inch diameter, 13%
have 3-8 inch diameter, and the rest have over 8 inch diameter. What will the distribution of diameters be in 12 years?
```{r}
inch_0_1 <- 0.50
inch_1_3 <- 0.28
inch_3_8 <- 0.13
inch_8 <- 1-(inch_0_1+inch_1_3+inch_3_8)
current <- c(inch_0_1,inch_1_3,inch_3_8,inch_8)

P_12 <- matrix.power(transition_matrix_2,4)
current%*%P_12

```
Problem 3: An electronic device contains a number of interchangeable parts that operate independently. Each component
is subject to random failure which renders it completely inoperable. All components are inspected at the end of every week.A component which is found to have failed is replaced with an identical new component. Based on historical data, a new component has 0.2 probability of failing in the first week. Components that are 1 week old have a 0.375 probability of failingin the second week. Components that are 2 weeks old have a 0.8 probability of failing in the third week. No components lastonger than 4 weeks. Let the process be represented by the following state variable:
0:0.2 probability of failing 
1:0.375 probability of failing
2:0.8 probability of failing
3:0  probability of failing

a.What is the transition matrix for this process?
```{r}
week_0<- c(0.2,0.8,0,0)
week_1<- c(0.375,0,0.625,0) 
week_2<- c(0.8,0,0,0.2) 
week_3<- c(1,0,0,0) 

x<- cbind(week_0,week_1,week_2,week_3)
colnames(x)<- c("week_0","week_1","week_2","week_3")
rownames(x)<- c("week_0","week_1","week_2","week_3")
transition_matrix_3<- matrix(x, nrow =4 ,byrow=TRUE)
dtmcA <- new("markovchain",transitionMatrix=transition_matrix_3,states=c("week_0","week_1","week_2","week_3"), name="MarkovChain A") 
dtmcA
plot(dtmcA)
```

b. Determine E(3). What is the value of e(3)02 and what does it mean in the context of this problem? 
```{r}
P_2 <- matrix.power(transition_matrix_3,2)
P_3 <-  matrix.power(transition_matrix_3,3)

E <- diag(4) + transition_matrix_3 + P_2 + P_3 
E[1,3]

```
c.If a device contains a component that is two weeks old, what is the probability that the next time that it has a component
that is two weeks old is 5 weeks from now?
```{r}
P_5 <- matrix.power(transition_matrix_3,5)
P_5[3,3]
```
What is the mean recurrence time for components being new? Find your answer by taking the reciprocal of the steady
state probability.
```{r}
library(MASS)
SS <- t(transition_matrix_3) - diag(4) 
SS<- rbind(SS[1:3,],c(1,1,1,1))
ginv(SS)%*%(c(0,0,0,1))
```

